---
title: "Duo et al 2018 Zheng8eq clusters"
author: "Will Townes"
output: html_document
---

assumes 01_filter_genes.Rmd has been run and informative gene ranks are stored in data/genefilter_ranks_8eq.txt.

```{r}
library(tidyverse); theme_set(theme_bw())
suppressPackageStartupMessages(library(SingleCellExperiment))
library(mclust)
#library(Seurat)
library(DuoClustering2018)
source("./real/util/functions_dimreduce.R") #loads glmpca, existing, and util/functions
ari<-mclust::adjustedRandIndex
fp<-file.path
bp<-"./real/duo_2018"
pth<-fp(bp,"results/fig")
if(!dir.exists(pth)){
  dir.create(pth,recursive=TRUE)
}
sp<-TRUE #save plots?
if(!dir.exists(fp(bp,"data"))) dir.create(fp(bp,"data"))
if(!dir.exists(fp(bp,"embeddings"))) dir.create(fp(bp,"embeddings"))
```

## Data Loading

```{r}
J<-2000
sce<-sce_full_Zhengmix8eq()
rk<-read.table(fp(bp,"data","genefilter_ranks_8eq.txt"))
sce_d<-sce[rownames(rk)[rk$devb<=J],]
sce_h<-sce[rownames(rk)[rk$hvg<=J],]
```

## Dimension reductions

```{r}
L<-15
res_d<-dimreduce_all(counts(sce_d),L)
write.table(res_d$factors,fp(bp,"embeddings/zheng8eq_dev_G2000_L15.txt"),quote=FALSE,row.names=FALSE)
res_h<-dimreduce_all(counts(sce_h),L)
write.table(res_h$factors,fp(bp,"embeddings/zheng8eq_hvg_G2000_L15.txt"),quote=FALSE,row.names=FALSE)

tt<-format_elapsed(res_d$elapsed,res_h$elapsed)
write.table(tt,fp(bp,"results/elapsed_zheng8eq_G2000_L15.txt"),quote=FALSE,row.names=FALSE)
```

### Make plots of dimension reductions

Load the embeddings and organize for plotting

```{r}
sce<-sce_full_Zhengmix8eq()
cm<-as.data.frame(colData(sce))
cm<-data.frame(cell_id=rownames(cm),celltype=cm$phenoid)
res_d<-read.table(fp(bp,"embeddings/zheng8eq_dev_G2000_L15.txt"),header=TRUE)
res_h<-read.table(fp(bp,"embeddings/zheng8eq_hvg_G2000_L15.txt"),header=TRUE)
res_d$genefilter<-"dev"
res_h$genefilter<-"hvg"
theirs<-c("pca_log","zinbwave")
pd<-rbind(subset(res_d,!(dimreduce %in% theirs)),subset(res_h,dimreduce %in% theirs))
pd<-merge(pd,cm)
#sanity check plot
ggplot(pd,aes(x=dim1,y=dim2,colour=celltype))+geom_point()+facet_wrap(~dimreduce,scales="free")
```

Visualize with UMAP

```{r}
L<-15
uf<-function(dr,...){
  d<-subset(pd,dimreduce==dr)
  cn<-paste0("dim",1:L)
  factors<-umap(t(d[,cn]),...)
  cbind(factors,d[,!(colnames(d) %in% cn)])
}
pd2<-do.call(rbind,lapply(unique(pd$dimreduce),uf,min_dist=1))
ggplot(pd2,aes(x=dim1,y=dim2,colour=celltype))+geom_point(size=.2) +facet_wrap(~dimreduce,scales="free") +guides(colour=guide_legend(override.aes=list(size=2)))
if(sp){ ggsave(fp(pth,"zheng8eq_umap.pdf"),width=7,height=4) }
```

### Duo clustering performance summaries

No method got ARI above 0.75

```{r}
sn<-"Zhengmix8eq"
for(vg in c("Expr10","HVG10")){
  n<-paste0("clustering_summary_filtered",vg,"_",sn,"_v2")
  m<-do.call(n,list())
  m_ari<-m %>% group_by(method,run,k,resolution) %>% summarise(ari=ari(cluster,trueclass)) %>% filter(!is.nan(ari))
  max(m_ari$ari)
  #top score for each method
  print(m_ari %>% group_by(method) %>% summarise(i=which.max(ari),run=run[i],k=k[i],resolution=resolution[i],ari=ari[i]) %>% arrange(desc(ari)))
}
```

### PCA

using log2(1+CPM)

```{r}
L<-5; K<-8
system.time(factors<-pca(Yl2,L))
sz<-colSums(Y)
pd<-cbind(factors,celltype=cm$phenoid,pz=colMeans(Y==0),z=log10(sz))
ggplot(pd,aes(x=dim1,y=dim2,colour=celltype))+geom_point()
if(sp) ggsave(fp(pth,"zheng8eq_log1p_pca12.pdf"),width=6,height=4)
ggplot(pd,aes(x=dim3,y=dim4,colour=celltype))+geom_point()
ggplot(pd,aes(x=pz,y=dim1,colour=celltype))+geom_point(show.legend = FALSE)+xlab("fraction of zeros")
if(sp) ggsave(fp(pth,"zheng8eq_log1p_pz_pc1.pdf"),width=4,height=4)
ggplot(pd,aes(x=z,y=pz,colour=celltype))+geom_point()
round(cor(pd[,colnames(pd)!="celltype"]),1)
cl<-kmeans(factors,K,nstart=25)$cl
table(pd$celltype,cl)
ari(cl,pd$celltype) #.41
mcl_res<-Mclust(factors,K)
ari(mcl_res$classification,pd$celltype) #.63
```

using deviance residuals

```{r}
L<-5; K<-8
system.time(pc0<-pca(Ydr,L,ret_obj=TRUE,center=TRUE,scale=TRUE))
factors<-pc0$factors; pc<-pc0$obj
sz<-colMeans(Y)
pd<-cbind(factors,celltype=cm$phenoid,pz=colMeans(Y==0),z=log10(sz))
ggplot(pd,aes(x=dim1,y=dim2,colour=celltype))+geom_point(show.legend=TRUE)
if(sp) ggsave(fp(pth,"zheng8eq_dr_pca12.pdf"),width=6,height=4)
ggplot(pd,aes(x=pz,y=dim1,colour=celltype))+geom_point(show.legend=FALSE)
if(sp) ggsave(fp(pth,"zheng8eq_dr_pz_pc1.pdf"),width=4,height=4)
kmeans_res<-kmeans(factors,K,nstart=25)
cl<-kmeans_res$cluster
ari(cl,pd$celltype) #.57
km_mu<-as.data.frame(kmeans_res$centers)
ggplot(pd,aes(x=dim1,y=dim2))+geom_point(aes(colour=celltype),size=.5)+geom_point(data=km_mu,shape="x",colour="black",size=5)+ggtitle("PCA on Fractions with k-means centers")

mcl_res<-Mclust(factors,K)
summary(mcl_res)
ari(mcl_res$classification,pd$celltype) #.69
mcl_mu<-as.data.frame(t(mcl_res$parameters$mean))
ggplot(pd,aes(x=dim1,y=dim2))+geom_point(aes(colour=celltype),size=.5)+geom_point(data=mcl_mu,shape="x",colour="black",size=5)+ggtitle("PCA on Fractions with k-means centers")
```

GLM-PCA: Poisson likelihood

```{r}
L<-5; K<-8
system.time(res<-glmpca(Y,L,fam="poi",verbose=TRUE)) #72 sec, 23 iter
plot(res$dev,type="l",log="y")
factors<-res$factors

sz<-colMeans(Y)
pd<-cbind(factors,celltype=cm$phenoid,pz=colMeans(Y==0),z=log10(sz))
ggplot(pd,aes(x=dim1,y=dim2,colour=celltype))+geom_point(show.legend=TRUE)
if(sp) ggsave(fp(pth,"zheng8eq_glmpca12.pdf"),width=6,height=4)
ggplot(pd,aes(x=dim3,y=dim4,colour=celltype))+geom_point(show.legend=FALSE)
ggplot(pd,aes(x=pz,y=dim1,colour=celltype))+geom_point(show.legend=FALSE)
if(sp) ggsave(fp(pth,"zheng8eq_pz_glmpca1.pdf"),width=4,height=4)
kmeans_res<-kmeans(factors,K,nstart=25)
cl<-kmeans_res$cluster
ari(cl,pd$celltype) #.74
mcl_res<-Mclust(factors,K)
ari(mcl_res$classification,pd$celltype) #.74
```

GLM-PCA: negative binomial likelihood

```{r}
L<-5; K<-8
system.time(res<-glmpca(Y,L,fam="nb",verbose=TRUE,nb_theta=100)) #56 iter, 6.4 min
plot(res$dev,type="l",log="y")
factors<-res$factors

sz<-colMeans(Y)
pd<-cbind(factors,celltype=cm$phenoid,pz=colMeans(Y==0),z=log10(sz))
ggplot(pd,aes(x=dim1,y=dim2,colour=celltype))+geom_point(show.legend=TRUE)
if(sp) ggsave(fp(pth,"zheng8eq_glmpca_nb12.pdf"),width=6,height=4)
ggplot(pd,aes(x=dim3,y=dim4,colour=celltype))+geom_point(show.legend=FALSE)
ggplot(pd,aes(x=pz,y=dim1,colour=celltype))+geom_point(show.legend=FALSE)
if(sp) ggsave(fp(pth,"zheng8eq_pz_glmpca_nb1.pdf"),width=4,height=4)
kmeans_res<-kmeans(factors,K,nstart=25)
cl<-kmeans_res$cluster
ari(cl,pd$celltype) #.68
mcl_res<-Mclust(factors,K)
ari(mcl_res$classification,pd$celltype) #.73
```

ZINB-WAVE

```{r}
L<-5; K<-8
system.time(factors<-zinbwave(Y,L)) #35.4 min

sz<-colMeans(Y)
pd<-cbind(factors,celltype=cm$phenoid,pz=colMeans(Y==0),z=log10(sz))
ggplot(pd,aes(x=dim1,y=dim2,colour=celltype))+geom_point(show.legend=TRUE)
if(sp) ggsave(fp(pth,"zheng8eq_zinbwave12.pdf"),width=6,height=4)
ggplot(pd,aes(x=dim3,y=dim4,colour=celltype))+geom_point(show.legend=FALSE)
ggplot(pd,aes(x=pz,y=dim1,colour=celltype))+geom_point(show.legend=FALSE)
if(sp) ggsave(fp(pth,"zheng8eq_pz_zinbwave1.pdf"),width=4,height=4)
kmeans_res<-kmeans(factors,K,nstart=25)
cl<-kmeans_res$cluster
ari(cl,pd$celltype) #.64
mcl_res<-Mclust(factors,K)
ari(mcl_res$classification,pd$celltype) #.67
```

Focusing only on hard-to-cluster T cells

```{r}
L<-4; K<-2
hard<-which(cm$phenoid %in% c("cd4.t.helper","regulatory.t"))
#ggplot(pd[hard,],aes(x=dim1,y=dim2,colour=cl))+geom_point(size=1,show.legend=TRUE)
Yhard<-Y[,hard]
Yhard<-Yhard[rowSums(Yhard>0)>0,]
res<-glmpca(Yhard,L,verbose=TRUE)#,init=list(factors=pc0$factors))
plot(log(res$dev),type="l")
factors<-res$factors
rownames(factors)<-colnames(Yhard)

sz<-colMeans(Yhard)
pd<-cbind(factors,celltype=cm$phenoid[hard],pz=colMeans(Yhard==0),z=log10(sz))
ggplot(pd,aes(x=dim1,y=dim2,colour=celltype))+geom_point(size=1,show.legend=TRUE)
ggplot(pd,aes(x=dim3,y=dim4,colour=celltype))+geom_point(size=.5,show.legend=FALSE)

gpc<-as.matrix(factors)
seu<-CreateSeuratObject(raw.data=Yhard)
seu<-SetDimReduction(seu,reduction.type="gpc",slot="cell.embeddings",new.data=gpc)
seu<-SetDimReduction(seu,reduction.type="gpc",slot="key",new.data="dim")
seu<-FindClusters(seu,reduction.type="gpc",dims.use=1:L,resolution=0.1,print.output=0)
s<-FetchData(seu,"ident")[[1]]
ari(s,cm$phenoid[hard]) #.78 best with poi regress out nUMI
table(cm$phenoid[hard],s)
```

## Old Stuff

### Non-Symmetric Correspondence Analysis

```{r}
nsca<-function(Xtable,k){
  #modified from nscabasic in CAvariants
  X <- Xtable/sum(Xtable)
  I <- nrow(X)
  J <- ncol(X)
  Imass <- rowSums(X)
  tauden <- 1 - sum(Imass^2)
  rsums <- as.vector(rowSums(X))
  csums <- as.vector(colSums(X))
  dcm1 <- 1/(csums + (csums == 0)) * (1 - (csums == 0))
  dcmh <- sqrt(dcm1)
  Z <- t(t(X - rsums %*% t(csums)) * dcmh)
  Y <- svd(Z, nu = k, nv = k)
  factors <- dcmh * Y$v #t(t(Y$v)*Y$d[1:k])
  colnames(factors)<-paste0("dim",1:k)
  as.data.frame(factors)
}

factors<-nsca(Y,8)
pd<-cbind(factors,celltype=cm$phenoid,pz=colMeans(Y==0),z=log10(sz))
ggplot(pd,aes(x=dim1,y=dim2,colour=celltype))+geom_point(size=.5)
ggplot(pd,aes(x=dim3,y=dim4,colour=celltype))+geom_point(size=.5)
ggplot(pd,aes(x=dim5,y=dim6,colour=celltype))+geom_point(size=.5)
cl<-kmeans(factors[,1:6],8,nstart=25)$cl
ari(cl,pd$celltype)
mcl_res<-Mclust(factors,8)
summary(mcl_res)
ari(mcl_res$classification,pd$celltype)
```

geometric NMF

```{r}
source("./algs/nmf.R")
fclustsim(U,pd$cl)
U2<-t(apply(U,1,function(u){if(max(u)>.5){ round(u) } else { u }}))
fclustsim(U2,pd$cl)
cl1<-apply(U,1,which.max)
barplot(sort(table(cl1),decreasing=TRUE),xlab="cluster ID",ylab="number of cells")
fclustsim(cl1,pd$cl)
ari(cl1,pd$cl)
table(cl1,pd$cl)
#plot the prototype vectors in PC space
pc<-pca(Yfrac,4,ret_obj=TRUE)
factors<-pc$factors; pc<-pc$obj
vv<-as.data.frame(nmf2pca(pc,t(t(V)/colSums(V))))
colnames(vv)<-paste0("dim",1:ncol(vv))
ggplot(pd,aes(x=dim1,y=dim2))+geom_point(aes(colour=cl),size=.5)+geom_text(data=vv,aes(label=rownames(vv)),colour="black",size=5)+ggtitle("NMF prototypes in PCA space")
ggplot(pd,aes(x=dim3,y=dim4))+geom_point(aes(colour=cl),size=.5)+geom_text(data=vv,aes(label=rownames(vv)),colour="black",size=5)+ggtitle("NMF prototypes in PCA space")
pd2<-as.data.frame(U)
colnames(pd2)<-paste0("dim",1:ncol(U))
pd2$cl<-pd$cl
ggplot(pd2,aes(x=dim1,y=dim2,col=cl))+geom_point()
ggplot(pd2,aes(x=dim3,y=dim4,col=cl))+geom_point()

cl2<-kmeans(U,4)$cl
fclustsim(cl2,pd$cl)
ari(cl2,pd$cl)
cl3<-apply(U[,1:4],1,which.max)
fclustsim(cl3,pd$cl)
table(cl3,pd$cl)
```

Try ILR transform with imputed zero counts based on Martin-Fernandez

```{r}
det_rate<-colMeans(Yfrac==0)
G<-nrow(Yfrac)
a0<-G #dirichlet prior concentration param
Yfrac2<-t(t(Yfrac)*(1-det_rate*a0/(a0+sz))) #adjust positive values down
Z<-Yfrac2==0
for(n in 1:ncol(Yfrac)){
  Yfrac2[Z[,n],n]<-a0/(G*(sz[n]+a0)) #impute zero values
}
X<-t(compositions::ilr(t(Yfrac2)))
factors<-pca(X,10)
pd<-cbind(factors,cl=cm$level1class,pz=colMeans(Y==0),z=log10(sz))
ggplot(pd,aes(x=dim1,y=dim2,colour=cl))+geom_point(size=.5)
ggplot(pd,aes(x=dim3,y=dim4,colour=cl))+geom_point(size=.5)
ggplot(pd,aes(x=dim5,y=dim6,colour=cl))+geom_point(size=.5)
ggplot(pd,aes(x=pz,y=dim1,colour=cl))+geom_point(size=.5)
round(cor(pd[,colnames(pd)!="cl"]),1)
cl<-kmeans(factors,7)$cl
table(pd$cl,cl)
fclustsim(cl,pd$cl)
cl2<-cutree(hclust(angular_distance(factors),"ave"),7)
table(pd$cl,cl2)
fclustsim(cl2,pd$cl)
```

# Session information

```{r}
sessionInfo()
```