---
title: "Zheng Monocytes Exploratory"
author: "Will Townes"
date: "12/04/2018"
output: html_document
---

Zheng 2017 Monocytes (10x)

```{r}
library(SingleCellExperiment)
library(cowplot)
#library(ggplot2)
library(Matrix)
source("./util/functions.R")
source("./algs/existing.R")
source("./util/clustering.R")
source("./algs/glmpca.R")
fp<-file.path
bp<-"./real/zheng_2017_monocytes"
pth<-fp(bp,"results/fig")
if(!dir.exists(pth)){
  dir.create(pth,recursive=TRUE)
}
sp<-TRUE #save plots?
ggs<-function(plt,w=6,h=4,...){
  if(sp){ ggsave(file=fp(pth,plt),width=w,height=h,...) }
}
```

Load UMI counts matrix and read counts matrix

```{r}
sce<-readRDS(fp(bp,"data/01_sce_all_genes_all_cells.rds"))
m<-assay(sce,"counts")
Y<-as.matrix(m)
sz<-colSums(Y)
Ycpm<-1e6*t(t(Y)/sz)
Yl2<-log2(1+Ycpm)
z<-log10(sz)
pz<-colMeans(Y==0)
Ypr<-null_residuals(Y,type="pearson")
Ydr<-null_residuals(Y,type="deviance")
rc<-assay(sce,"read_counts")
Yl2scran<-assay(sce,"logcounts")

set.seed(205) #reproducibility
ss<-sample.int(nrow(Y),size=5000) #subset random 5000 genes
```

plot histogram of a particular gene for reads and umis

```{r}
rmx<-apply(Y,1,max)
g<-match(20,rmx)
gn<-rownames(Y)[g]
pd<-list(data.frame(prot="raw_umi_counts",expr=Y[g,]))
pd[[2]]<-data.frame(prot="raw_read_counts",expr=rc[g,])
pd<-do.call(rbind,pd)
plts<-list()
ymax<-750
(plts[[1]]<-ggplot(subset(pd,prot=="raw_umi_counts"),aes(x=expr))+geom_bar()+ylim(NA,ymax)+theme_bw()+xlab(paste(gn,"UMI counts"))+ylab("number of droplets in bin"))
ggs("monocytes_umi_distr.pdf")
(plts[[2]]<-ggplot(subset(pd,prot=="raw_read_counts"),aes(x=expr))+geom_histogram(bins=100)+ylim(NA,ymax)+theme_bw()+xlab(paste(gn,"read counts"))+ylab("number of droplets in bin"))#+ylab(NULL))
ggs("monocytes_read_distr.pdf")
plot_grid(plotlist=plts)
ggs("reads_vs_umis_monocytes.pdf")
```

Fraction of zeros in each droplet determined by total UMIs.

```{r}
pd<-data.frame(sz=sz,pz=pz)
(plt<-ggplot(pd,aes(x=sz,y=pz))+geom_point()+theme_bw()+xlab("total UMI per droplet")+ylab("fraction of zeros per droplet")+scale_x_log10())
(plt2<-ggExtra::ggMarginal(plt,type="histogram",fill="white",bins=100))
ggs("total_umi_vs_pzero_monocytes.pdf",plot=plt2)
```

Log transformation distorts distribution

```{r}
#find gene where max count across all droplets was 5
rmx<-apply(Y,1,max)
g<-match(20,rmx)
gn<-rownames(Y)[g] #ENSG00000114391
plts<-list()
(plts[["umi"]]<-ggplot(data.frame(x=Y[g,]),aes(x=x))+geom_bar()+theme_bw()+xlab("counts")+ylab(NULL))
ggs("monocytes_distr_a_umi_counts.pdf",w=3,h=2)
(plts[["cpm"]]<-ggplot(data.frame(x=Ycpm[g,]),aes(x=x))+geom_histogram(bins=100)+theme_bw()+xlab("CPM")+ylab(NULL))
ggs("monocytes_distr_b_cpm.pdf",w=3,h=2)
(plts[["logcpm"]]<-ggplot(data.frame(x=Yl2[g,]),aes(x=x))+geom_histogram(bins=100)+theme_bw()+xlab("log2(1+CPM)")+ylab(NULL))
ggs("monocytes_distr_c_log2cpm.pdf",w=3,h=2)
(plts[["logcptt"]]<-ggplot(data.frame(x=log2(1+Ycpm[g,]/100)),aes(x=x))+geom_histogram(bins=100)+theme_bw()+xlab("log2(1+CPTT)")+ylab(NULL))
ggs("monocytes_distr_d_log2cptt.pdf",w=3,h=2)
(plts[["logscran"]]<-ggplot(data.frame(x=Yl2scran[g,]),aes(x=x))+geom_histogram(bins=100)+theme_bw()+xlab("log2(1+scran)")+ylab(NULL))
ggs("monocytes_distr_e_log2scran.pdf",w=3,h=2)

plot_grid(plotlist=plts,nrow=2)
ggs("counts_vs_log_cpm_histograms_monocytes.pdf",w=7.5,h=5)
```

Library Size/ Fraction of Zeros dominates first PC even after normalization/ log transform

```{r}
factors<-pca(Yl2[ss,],10)
pd<-cbind(factors[,1:3],zero_frac=pz,log_total_umi=z)
if(cor(pd$zero_frac,pd$dim1)<0) pd$dim1<- -pd$dim1
ggplot(pd,aes(x=dim1,y=dim2,colour=zero_frac))+geom_point()+scale_color_continuous(low="blue",high="red")
ggs("log_cpm_pca12_zerofrac_monocytes.pdf")
ggplot(pd,aes(x=zero_frac,y=dim1,colour=log_total_umi))+geom_point()+scale_color_continuous(low="red",high="blue")+xlab("zero fraction")+ylab("dimension 1")
ggs("log_cpm_pca1_zerofrac_total_umi_monocytes.pdf")
round(cor(pd[,colnames(pd)!="cl"]),1)

#apply clustering to the result of PCA
cl<-seurat_cluster(factors,res=0.5)
pd<-cbind(cl,zero_frac=pz)
ggplot(pd,aes(x=factor(cluster),y=zero_frac))+geom_boxplot()+xlab("Seurat cluster ID")+ylab("zero fraction")
ggs("pca_seurat_cluster_zero_frac_monocytes.pdf")

#tsne on top 50 PCs
tsf<-tsne(Yl2[ss,],2,perplexity=30,pca=TRUE)
pd2<-cbind(tsf,zero_frac=pz,log_total_umi=z)
#nonlinear color scale from https://stackoverflow.com/a/12838299
qn<-quantile(pd2$zero_frac, seq(from=0,to=1,length.out=20))
cols<-colorRampPalette(c("blue","purple","red"))(20)
vals<-scales::rescale(qn) 
ggplot(pd2,aes(x=dim1,y=dim2,colour=zero_frac))+geom_point(size=1)+scale_color_gradientn(colours=cols,values=vals)+xlab("dimension 1")+ylab("dimension 2")
ggs("log_cpm_tsne12_zerofrac_monocytes.pdf")
```

same procedure as above except using GLM-PCA

```{r}
ff<-fp(bp,"results","glmpca_L10_rand_5K_genes.txt")
if(!file.exists(ff)){
  ctl<-list(maxIter=500,eps=1e-4)
  toc<-system.time(res<-glmpca(Y[ss,],10,verbose=TRUE,ctl=ctl))
  factors<-res$factors
  write.table(factors,file=ff)
} else {
  factors<-read.table(ff,header=TRUE)
}
pd<-cbind(factors,zero_frac=pz,log_total_umi=z)
if(cor(pd$zero_frac,pd$dim1)<0) pd$dim1<- -pd$dim1
ggplot(pd,aes(x=dim1,y=dim2,colour=zero_frac))+geom_point()+scale_color_continuous(low="blue",high="red")
ggs("glmpca12_zerofrac_monocytes.pdf")
ggplot(pd,aes(x=zero_frac,y=dim1,colour=log_total_umi))+geom_point()+scale_color_continuous(low="red",high="blue")+xlab("zero fraction")+ylab("dimension 1")
ggs("glmpca1_zerofrac_total_umi_monocytes.pdf")
round(cor(pd[,colnames(pd)!="cl"]),1)

#apply clustering to the result of GLMPCA
cl<-seurat_cluster(factors,res=0.5)
pd<-cbind(cl,zero_frac=pz)
ggplot(pd,aes(x=factor(cluster),y=zero_frac))+geom_boxplot()+xlab("Seurat cluster ID")+ylab("zero fraction")
ggs("glmpca_seurat_cluster_zero_frac_monocytes.pdf")

#tsne on top 50 GLMPCA factors
ff<-fp(bp,"results","glmpca_L50_rand_5K_genes.txt")
if(!file.exists(ff)){
  ctl<-list(maxIter=500,eps=1e-4)
  toc<-system.time(res<-glmpca(Y[ss,],50,verbose=TRUE,ctl=ctl))
  factors<-res$factors
  write.table(factors,file=ff)
} else {
  factors<-read.table(ff,header=TRUE)
}
tsf<-tsne(t(factors),2,perplexity=30,pca=FALSE,rmzero=FALSE)
pd2<-cbind(tsf,zero_frac=pz,log_total_umi=z)
#nonlinear color scale from https://stackoverflow.com/a/12838299
qn<-quantile(pd2$zero_frac, seq(from=0,to=1,length.out=20))
cols<-colorRampPalette(c("blue","purple","red"))(20)
vals<-scales::rescale(qn) 
ggplot(pd2,aes(x=dim1,y=dim2,colour=zero_frac))+geom_point(size=1)+scale_color_gradientn(colours=cols,values=vals)+xlab("dimension 1")+ylab("dimension 2")
ggs("glmpca_tsne12_zerofrac_monocytes.pdf")
```

more tsne combinations

```{r}
plx<-c(1,2,5,10,30,100)
Ypca<-t(pca(Yl2,50))
f<-function(p){
  factors<-tsne(Ypca,2,perplexity=p,pca=FALSE)
  cbind(factors,zero_frac=pz,log_total_umi=z,perplexity=p)
}
pd<-do.call(rbind,lapply(plx,f))
ggplot(pd,aes(x=dim1,y=dim2,colour=zero_frac))+geom_point(size=1)+scale_color_continuous(low="blue",high="red")+facet_wrap(~perplexity,scales="free",nrow=3)
ggs("log_cpm_tsne12_zerofrac_monocytes.pdf",w=10,h=15)

#larger number of perplexities
plx2<-c(.5,1:20,40,60,80,100,150,200)
pd2<-do.call(rbind,lapply(plx2,f))
pdf(fp(pth,"log_cpm_tsne12_zerofrac_monocytes_big.pdf"))
for(p in plx2){
  print(ggplot(subset(pd2,perplexity==p),aes(x=dim1,y=dim2,colour=zero_frac))+geom_point()+scale_color_continuous(low="blue",high="red")+ggtitle(paste("perplexity=",p)))
}
dev.off()
```

Show relationship between mean expression and probability of zero follows multinomial prediction.

```{r}
Down_Sample_Matrix<-function(expr_mat){
  #adapted from https://hemberg-lab.github.io/scRNA.seq.course/cleaning-the-expression-matrix.html#normalisations
  min_lib_size<-min(colSums(expr_mat))
  down_sample<-function(x){
    prob <- min_lib_size/sum(x)
    unlist(lapply(x,function(y){rbinom(1, y, prob)}))
  }
  apply(expr_mat, 2, down_sample)
}

#downsample to normalize droplet size (total UMI)
Yds<-Down_Sample_Matrix(Y)
Yds<-Yds[rowSums(Yds)>0,]
#variance=mean, suggests poisson
m<-rowMeans(Yds); v<-apply(Yds,1,var)
summary(v/m)
plot(log(m),log(v),xlab="log(mean)",ylab="log(var)")
abline(0,1,col="blue") #poi
curve(x+log1p(exp(x)/5),from=-8,to=3,add=TRUE,lty=2,col="red") #nb

N<-median(colSums(Yds))
predict_zeros_binom<-function(x){(1-exp(x)/N)^N} #binomial
predict_zeros_poi<-function(x){exp(-exp(x))}
predict_zeros_nb<-function(x,phi=2){
  exp(-phi*log1p(exp(x-log(phi))))
}
pd<-data.frame(log_mean=log(m),frac_zero=rowMeans(Yds==0))
xlo<-min(pd$log_mean)
xhi<-max(pd$log_mean)
xcv<-data.frame(x=c(xlo,xhi))
ggplot(xcv)+geom_point(data=pd,aes(x=log_mean,y=frac_zero),alpha=.5) +stat_function(aes(x,color="bin"),fun=predict_zeros_binom) +stat_function(aes(x,color="poi"),fun=predict_zeros_poi) +stat_function(aes(x,color="nb"),fun=predict_zeros_nb) #+scale_color_manual("model",breaks=c("bin","poi","nb"),values=c("blue","green","red"))
#ggs("logmean_pzero_binom_monocytes.pdf")

pdf(fp(pth,"logmean_pzero_monocytes.pdf"),width=6,height=4)
#same plot but using base plot
with(pd,plot(log_mean,frac_zero,xlab="log of mean expression",ylab="fraction of zero droplets",cex=1.5))
curve(predict_zeros_binom,from=xlo,to=xhi,col="blue",lwd=4,add=TRUE)
curve(predict_zeros_poi,from=xlo,to=xhi,col="green",lwd=3,lty=2,add=TRUE)
curve(predict_zeros_nb(x,phi=4),from=xlo,to=xhi,col="red",lwd=3,lty=3,add=TRUE)
legend("bottomleft",c("Multinomial","Poisson","Negative Binomial"),lty=c(1,2,3),lwd=c(4,3,3),col=c("blue","green","red"))
dev.off()
```

testing for zero inflation

```{r}
score_test<-function(x,m){
  #x a vector of count data
  #m a vector of multiplicative offsets (log(m) is the offset in glm terms)
  #computes a score test statistic for zero inflation from (van den Broek 1995)
  #test statistic is distributed as chisquare(1). 
  #Rejection of null means simple poisson insufficient, may be zero inflated/ overdispersed
  #model is x[i]~poisson(m[i]*lambda)
  lambda<-sum(x)/sum(m) #Poisson MLE, scalar
  mu<-lambda*m #Poisson mean, adjusted for offset, vector
  p0<-exp(-mu) #predicted probability of zero, vector
  z<- x==0
  num<-sum((z-p0)/p0)^2
  denom<-sum((1-p0)/p0)-sum(mu)
  num/denom #test statistic as chisq(1), high value=more zero inflation
}

#hypothesis testing of whether there is zero inflation
thresh<-qchisq(.05/nrow(Yds),1,lower.tail=FALSE)
s<-apply(Yds,1,score_test,colMeans(Yds))
hist(s)
abline(v=thresh,lty=2)
table(s>thresh)

pd$zero_inflated<-ifelse(s>thresh,"zero_inflated","normal")
#predict_zeros<-function(x){exp(-exp(x))} #poisson
ggplot(xcv)+xlab("log of mean expression")+ylab("fraction of zero droplets")+geom_point(data=pd,size=1,aes(x=log_mean,y=frac_zero,colour=zero_inflated))+scale_color_manual(breaks=c("normal","zero_inflated"),values=c("blue","red"))+stat_function(aes(x),fun=predict_zeros_binom)
```

```{r}
sessionInfo()
```
