---
title: "Zheng monocytes Model Selection"
author: "Will Townes"
output: html_document
---

Zheng 2017 Monocytes (10x)

```{r}
suppressPackageStartupMessages(library(SingleCellExperiment))
library(tidyverse); theme_set(theme_bw())
source("./util/functions.R")

fp<-file.path
bp<-"./real/zheng_2017_monocytes"
pth<-fp(bp,"results/fig")
if(!dir.exists(pth)){
  dir.create(pth,recursive=TRUE)
}
sp<-TRUE #save plots?
ggs<-function(plt,w=6,h=4,...){
  if(sp){ ggsave(file=fp(pth,plt),width=w,height=h,...) }
}

get_qc_covars<-function(m){
  #m is a count matrix (either UMI or reads)
  #rownames must be human ENSEMBL gene IDs
  #colnames are cell IDs
  #returns a data frame with QC metrics per cell
  #(total count, fraction zeros, frac mitochondria, cell cycle)
  #this data frame can be used as covariates to regress out
  gm<-ensembl2symbol(rownames(m),"hsapiens") #from util/functions.R
  mtg<-grepl("^MT-",gm$hgnc_symbol)
  sg<-gm$hgnc_symbol %in% Seurat::cc.genes$s.genes
  g2mg<-gm$hgnc_symbol %in% Seurat::cc.genes$g2m.genes
  
  cm<-data.frame(total_count=colSums(m),zero_frac=1-colMeans(m>0))
  cm$pct_mito<-colSums(m[mtg,])/cm$total_count
  cm$pct_cc_s_phase<-colSums(m[sg,])/cm$total_count
  cm$pct_cc_g2m_phase<-colSums(m[g2mg,])/cm$total_count
  cm
}
```

Load UMI counts matrix and read counts matrix. Compute QC metrics based on UMIs and read counts. Total counts, fraction of zeros, percentage mitochondrial, and cell cycle phase scores are the QC metrics we will try to regress out.

```{r}
sce0<-readRDS(fp(bp,"data/01_sce_all_genes_all_cells.rds"))
Xumi0<-get_qc_covars(as.matrix(assay(sce0,"counts")))
Xrc0<-get_qc_covars(as.matrix(assay(sce0,"read_counts")))
```

Load UMI counts matrix and read counts matrix

```{r}
cm<-colData(sce0)
set.seed(205) #reproducibility
ssc<-sample.int(ncol(sce0),size=500) #subset random cells
ssg<-sample.int(nrow(sce0),size=2000) #subset random genes
sce<-sce0[ssg,ssc]
#exclude the total counts since this is computed separately as an offset.
Xumi<-cbind(intercept=1,scale(Xumi0[ssc,-1]))
Xrc<-cbind(intercept=1,scale(Xrc0[ssc,-1]))
#remove genes from data matrices that have no variation or no zeros.
m<-as.matrix(assay(sce,"counts"))
#sz<-colSums(m)
#z<-log10(sz)
#pz<-1-colMeans(m>0)
rowmins<-apply(m,1,min)
rowsd<-apply(m,1,sd)
m2<-m[rowmins==0 & rowsd>1e-12,]
rc<-as.matrix(assay(sce,"read_counts"))
rowmins<-apply(rc,1,min)
rowsd<-apply(rc,1,sd)
rc2<-rc[rowmins==0 & rowsd>1e-12,]
print(dim(m2)) #3357 genes x 57 cells
summary(rowMeans(m2)) 
```

### Compare model fit for different likelihoods

Likelihoods:
* multinomial
* Poisson
* Dirichlet-multinomial
* negative binomial
* normal
* zero inflated Poisson
* zero inflated lognormal

All likelihoods include column means as offset. Have to remove genes with no zero values in order to include zero inflated models in comparisons.

```{r}
system.time(res<-bic_all(m2))
barplot(sort(res),xlab="likelihood",ylab="BIC",main="UMI counts",log="y")
res_rc<-bic_all(rc2)
barplot(sort(res_rc),xlab="likelihood",ylab="BIC",main="read counts")
dd<-as.data.frame(res)
colnames(dd)<-"BIC"
dd$dat<-"umi_counts"
dd$lik<-rownames(dd)
dd2<-as.data.frame(res_rc)
colnames(dd2)<-"BIC"
dd2$dat<-"read_counts"
dd2$lik<-rownames(dd2)
dd<-rbind(dd,dd2)
rownames(dd)<-NULL
dd<-dd[,c(2,3,1)]

#rerun with covariates on UMI data
system.time(res1<-bic_all(m2,Xumi))
system.time(res2<-bic_all(rc2,Xrc,liks=c("poi","zip","ziln")))
dd$covar<-FALSE
dd1<-data.frame(dat="umi_counts",lik=names(res1),BIC=res1,covar=TRUE)
dd2<-data.frame(dat="read_counts",lik=names(res2),BIC=res2,covar=TRUE)
dd<-rbind(dd,dd1,dd2)
rownames(dd)<-NULL

write.table(dd,fp(bp,"results/gof_bic.txt"),quote=FALSE,row.names=FALSE)
```

The best performing likelihood is Dirichlet-multinomial for UMI counts followed by multinomial then Poisson.

```{r}
pd<-read.table(fp(bp,"results/gof_bic.txt"),header=TRUE)
levels(pd$lik)[levels(pd$lik)=="normal"]<-"nml"
plt_func<-function(d){
  d$lik<-factor(d$lik,levels=d$lik[order(d$BIC)])
  ggplot(d,aes(x=lik,y=BIC))+geom_point(size=3)+xlab("distribution")
}
plt_func(subset(pd,dat=="umi_counts" & lik!="nml" & !covar))
ggs("gof_bic_zheng_monocytes_umi.pdf",w=3,h=2)
plt_func(subset(pd,dat=="read_counts" & !covar))
ggs("gof_bic_zheng_monocytes_reads.pdf",w=3,h=2)
plt_func(subset(pd,dat=="umi_counts" & covar))
ggs("gof_bic_zheng_umi_covar.pdf",w=3,h=2)
plt_func(subset(pd,dat=="read_counts" & covar))
ggs("gof_bic_zheng_reads_covar.pdf",w=3,h=2)

#compare whether adding covariates improved the fit
plt_func2<-function(d){
  pd2<-d %>% group_by(lik) %>% summarise(BIC=mean(BIC)) %>% arrange(BIC)
  d$lik<-factor(d$lik,levels=pd2$lik)
  ggplot(d,aes(x=lik,y=BIC,colour=covar,shape=covar))+geom_point(size=3)+xlab("distribution")
}
plt_func2(subset(pd,dat=="umi_counts" & lik!="nml"))
ggs("gof_bic_zheng_umi_both.pdf")
plt_func2(subset(pd,dat=="read_counts"))
ggs("gof_bic_zheng_reads_both.pdf")
```
