---
title: "Zheng ERCC Exploratory"
author: "Will Townes"
date: "12/04/2018"
output: html_document
---

ERCC spike ins

```{r}
library(SingleCellExperiment)
library(cowplot)
#library(ggplot2)
library(Matrix)
source("./util/functions.R")
source("./algs/existing.R")
source("./algs/glmpca.R")
source("./util/clustering.R")
fp<-file.path
bp<-"./real/zheng_2017_ercc"
pth<-fp(bp,"results/fig")
if(!dir.exists(pth)){
  dir.create(pth,recursive=TRUE)
}
sp<-TRUE #save plots?
ggs<-function(plt,w=6,h=4,...){
  if(sp){ ggsave(file=fp(pth,plt),width=w,height=h,...) }
}
```

useful functions

```{r}
Down_Sample_Matrix<-function(expr_mat){
  #adapted from https://hemberg-lab.github.io/scRNA.seq.course/cleaning-the-expression-matrix.html#normalisations
  min_lib_size<-min(colSums(expr_mat))
  down_sample<-function(x){
    prob <- min_lib_size/sum(x)
    unlist(lapply(x,function(y){rbinom(1, y, prob)}))
  }
  apply(expr_mat, 2, down_sample)
}

score_test<-function(x,m){
  #x a vector of count data
  #m a vector of multiplicative offsets (log(m) is the offset in glm terms)
  #computes a score test statistic for zero inflation from (van den Broek 1995)
  #test statistic is distributed as chisquare(1). 
  #Rejection of null means simple poisson insufficient, may be zero inflated/ overdispersed
  #model is x[i]~poisson(m[i]*lambda)
  lambda<-sum(x)/sum(m) #Poisson MLE, scalar
  mu<-lambda*m #Poisson mean, adjusted for offset, vector
  p0<-exp(-mu) #predicted probability of zero, vector
  z<- x==0
  num<-sum((z-p0)/p0)^2
  denom<-sum((1-p0)/p0)-sum(mu)
  num/denom #test statistic as chisq(1), high value=more zero inflation
}

lr_test<-function(x,m){
  if(min(x)>0){
    return(0)
  }
  full<-pscl::zeroinfl(x ~ 1 | 1, dist="poisson",offset=log(m))
  reduced<-glm(x~1,family="poisson",offset=log(m))
  2*as.numeric(logLik(full)-logLik(reduced))
}
```

Load UMI counts matrix and read counts matrix

```{r}
sce<-readRDS(fp(bp,"data/01_sce_all_genes_all_cells.rds"))
m<-assay(sce,"counts")
Y<-as.matrix(m)
sz<-colSums(Y)
Ycpm<-1e6*t(t(Y)/sz)
Yl2<-log2(1+Ycpm)
z<-log10(sz)
pz<-colMeans(Y==0)
Ypr<-null_residuals(Y,type="pearson")
Ydr<-null_residuals(Y,type="deviance")
rc<-assay(sce,"read_counts")
```

plot histogram of a particular ERCC for reads and umis

```{r}
rmx<-apply(Y,1,max)
g<-match(20,rmx)
gn<-rownames(Y)[g]
pd<-list(data.frame(prot="raw_umi_counts",expr=Y[g,]))
pd[[2]]<-data.frame(prot="raw_read_counts",expr=rc[g,])
pd<-do.call(rbind,pd)
plts<-list()
ymax<-250
(plts[[1]]<-ggplot(subset(pd,prot=="raw_umi_counts"),aes(x=expr))+geom_bar()+ylim(NA,ymax)+theme_bw()+xlab(paste(gn,"UMI counts"))+ylab("number of droplets in bin"))
ggs("ercc_umi_distr.pdf")
(plts[[2]]<-ggplot(subset(pd,prot=="raw_read_counts"),aes(x=expr))+geom_histogram(bins=100)+ylim(NA,ymax)+theme_bw()+xlab(paste(gn,"read counts"))+ylab("number of droplets in bin"))#+ylab(NULL))
ggs("ercc_read_distr.pdf")
plot_grid(plotlist=plts)
ggs("reads_vs_umis_ercc.pdf")
```

Fraction of zeros in each droplet determined by total UMIs.

```{r}
#ggplot(data.frame(z=z,pz=pz),aes(x=z,y=pz))+geom_point()+theme_bw()+xlab("log10(total UMI per droplet)")+ylab("fraction of zeros per droplet")
#ggs("total_umi_vs_pzero_ercc.pdf")

pd<-data.frame(sz=sz,pz=pz)
(plt<-ggplot(pd,aes(x=sz,y=pz))+geom_point()+theme_bw()+xlab("total UMI per droplet")+ylab("fraction of zeros per droplet")+scale_x_log10())
(plt2<-ggExtra::ggMarginal(plt,type="histogram",fill="white",bins=100))
ggs("total_umi_vs_pzero_ercc.pdf",plot=plt2)
```

Log transformation distorts distribution

```{r}
#find gene where max count across all droplets was 20
rmx<-apply(Y,1,max)
g<-match(20,rmx)
gn<-rownames(Y)[g] #ERCC-00163
plts<-list()
(plts[["umi"]]<-ggplot(data.frame(x=Y[g,]),aes(x=x))+geom_bar()+theme_bw()+xlab("counts")+ylab(NULL))
(plts[["cpm"]]<-ggplot(data.frame(x=Ycpm[g,]),aes(x=x))+geom_histogram(bins=100)+theme_bw()+xlab("CPM")+ylab(NULL))
(plts[["logcpm"]]<-ggplot(data.frame(x=Yl2[g,]),aes(x=x))+geom_histogram(bins=100)+theme_bw()+xlab("log2(1+CPM)")+ylab(NULL))
(plts[["logcptt"]]<-ggplot(data.frame(x=log2(1+Ycpm[g,]/100)),aes(x=x))+geom_histogram(bins=100)+theme_bw()+xlab("log2(1+CPTT)")+ylab(NULL))
plot_grid(plotlist=plts,nrow=2)
ggs("counts_vs_log_cpm_histograms_ercc.pdf",w=7.5,h=5)
```

Library Size/ Fraction of Zeros dominates first PC even after normalization/ log transform

```{r}
factors<-pca(Yl2,10)
pd<-cbind(factors[,1:5],zero_frac=pz,log_total_umi=z)
if(cor(pd$zero_frac,pd$dim1)<0) pd$dim1<- -pd$dim1
ggplot(pd,aes(x=dim1,y=dim2,colour=zero_frac))+geom_point()+scale_color_continuous(low="blue",high="red")
ggs("log_cpm_pca12_zerofrac_ercc.pdf")
ggplot(pd,aes(x=zero_frac,y=dim1,colour=log_total_umi))+geom_point()+scale_color_continuous(low="red",high="blue")+xlab("zero fraction")+ylab("dimension 1")
ggs("log_cpm_pca1_zerofrac_total_umi_ercc.pdf")
round(cor(pd[,colnames(pd)!="cl"]),1)

#apply clustering to the result of PCA
cl<-seurat_cluster(factors,res=0.5)
pd<-cbind(cl,zero_frac=pz)
ggplot(pd,aes(x=factor(cluster),y=zero_frac))+geom_boxplot()+xlab("Seurat cluster ID")+ylab("zero fraction")
ggs("pca_seurat_cluster_zero_frac_ercc.pdf")

#tsne on top 50 PCs
tsf<-tsne(Yl2,2,perplexity=30,pca=TRUE)
pd2<-cbind(tsf,zero_frac=pz,log_total_umi=z)
#nonlinear color scale from https://stackoverflow.com/a/12838299
qn<-quantile(pd2$zero_frac, seq(from=0,to=1,length.out=20))
cols<-colorRampPalette(c("blue","purple","red"))(20)
vals<-scales::rescale(qn) 
ggplot(pd2,aes(x=dim1,y=dim2,colour=zero_frac))+geom_point(size=1)+scale_color_gradientn(colours=cols,values=vals)+xlab("dimension 1")+ylab("dimension 2")
ggs("log_cpm_tsne12_zerofrac_ercc.pdf")
```

same as above but using GLM-PCA

```{r}
factors<-glmpca(Y,10,verbose=TRUE)$factors
pd<-cbind(factors,zero_frac=pz,log_total_umi=z)
if(cor(pd$zero_frac,pd$dim1)<0) pd$dim1<- -pd$dim1
ggplot(pd,aes(x=dim1,y=dim2,colour=zero_frac))+geom_point()+scale_color_continuous(low="blue",high="red")
ggs("glmpca12_zerofrac_ercc.pdf")
ggplot(pd,aes(x=zero_frac,y=dim1,colour=log_total_umi))+geom_point()+scale_color_continuous(low="red",high="blue")+xlab("zero fraction")+ylab("dimension 1")
ggs("glmpca1_zerofrac_total_umi_ercc.pdf")
round(cor(pd[,colnames(pd)!="cl"]),1)

#apply clustering to the result of GLMPCA
cl<-seurat_cluster(factors,res=0.5)
pd<-cbind(cl,zero_frac=pz)
ggplot(pd,aes(x=factor(cluster),y=zero_frac))+geom_boxplot()+xlab("Seurat cluster ID")+ylab("zero fraction")
ggs("glmpca_seurat_cluster_zero_frac_ercc.pdf")

#tsne on top 50 GLMPCA factors
ctl<-list(maxIter=500,eps=1e-4)
toc<-system.time(res<-glmpca(Y,50,verbose=TRUE,ctl=ctl)) #5mins
factors<-res$factors
tsf<-tsne(t(factors),2,perplexity=30,pca=FALSE,rmzero=FALSE)
pd2<-cbind(tsf,zero_frac=pz,log_total_umi=z)
#nonlinear color scale from https://stackoverflow.com/a/12838299
qn<-quantile(pd2$zero_frac, seq(from=0,to=1,length.out=20))
cols<-colorRampPalette(c("blue","purple","red"))(20)
vals<-scales::rescale(qn) 
ggplot(pd2,aes(x=dim1,y=dim2,colour=zero_frac))+geom_point(size=1)+scale_color_gradientn(colours=cols,values=vals)+xlab("dimension 1")+ylab("dimension 2")
ggs("glmpca_tsne12_zerofrac_ercc.pdf")
```

compare clustering of log-CPM pca vs glmpca

```{r}
silfunc<-function(f,cl,m="euclidean"){
  cl<-as.integer(cl$cluster)
  d<-dist(f,method=m)
  sil<-cluster::silhouette(cl,d)
  summary(sil)$avg.width
}
L<-2
f1<-pca(Yl2,L)
f2<-glmpca(Y,L)$factors
resolution<-0.08
c1<-seurat_cluster(f1,res=resolution)
c2<-seurat_cluster(f2,res=resolution)
silfunc(f1,c1)
silfunc(f2,c2)
table(c1$cluster)
table(c2$cluster)

k<-2
c1<-kmeans_cluster(f1,k=k)
c2<-kmeans_cluster(f2,k=k)
silfunc(f1,c1)
silfunc(f2,c2)
```

larger number of perplexities

```{r}
#plx<-c(1,2,5,10,30,100)
f<-function(p){
  factors<-tsne(Yl2,2,perplexity=p,pca=FALSE)
  cbind(factors,zero_frac=pz,log_total_umi=z,perplexity=p)
}
#pd<-do.call(rbind,lapply(plx,f))
#ggplot(pd,aes(x=dim1,y=dim2,colour=zero_frac))+geom_point(size=1)+scale_color_continuous(low="blue",high="red")+facet_wrap(~perplexity,scales="free",nrow=3)
#ggs("log_cpm_tsne12_zerofrac_ercc.pdf",w=10,h=15)

#larger number of perplexities
plx2<-c(.5,1:20,40,60,80,100,150,200)
pd2<-do.call(rbind,lapply(plx2,f))
pdf(fp(pth,"log_cpm_tsne12_zerofrac_ercc_big.pdf"))
for(p in plx2){
  print(ggplot(subset(pd2,perplexity==p),aes(x=dim1,y=dim2,colour=zero_frac))+geom_point()+scale_color_continuous(low="blue",high="red")+ggtitle(paste("perplexity=",p)))
}
dev.off()
```

Check for zero inflation. We fit a score test and a likelihood ratio test to assess whether there is zero inflation. In both cases, the null hypothesis is that the Poisson model is sufficient. For the likelihood ratio test, the alternative model is the zero inflated Poisson.

```{r}
#downsample to normalize droplet size (total UMI)
Yds<-Down_Sample_Matrix(Y[,colSums(Y)>10000])
#variance=mean, suggests poisson
m<-rowMeans(Yds); v<-apply(Yds,1,var)
summary(v/m)
plot(log(m),log(v),xlab="log(mean)",ylab="log(var)")
abline(0,1)

N<-median(colSums(Yds))
predict_zeros_binom<-function(x){(1-exp(x)/N)^N} #binomial
predict_zeros_poi<-function(x){exp(-exp(x))}
predict_zeros_nb<-function(x,phi=2){
  exp(-phi*log1p(exp(x-log(phi))))
}
pd<-data.frame(log_mean=log(m),frac_zero=rowMeans(Yds==0))
xlo<-min(pd$log_mean[pd$log_mean>-Inf])
xhi<-max(pd$log_mean)
xcv<-data.frame(x=c(xlo,xhi))
ggplot(xcv)+geom_point(data=pd,aes(x=log_mean,y=frac_zero),alpha=.5) +stat_function(aes(x,color="bin"),fun=predict_zeros_binom) +stat_function(aes(x,color="poi"),fun=predict_zeros_poi) #+scale_color_manual("model",breaks=c("bin","poi","nb"),values=c("blue","green","red"))

pdf(fp(pth,"logmean_pzero_ercc.pdf"),width=6,height=4)
#same plot but using base plot
with(pd,plot(log_mean,frac_zero,xlab="log of mean expression",ylab="fraction of zero droplets",cex=2))
curve(predict_zeros_binom,from=xlo,to=xhi,col="blue",lwd=5,add=TRUE)
curve(predict_zeros_poi,from=xlo,to=xhi,col="green",lwd=4,lty=2,add=TRUE)
legend("bottomleft",c("Multinomial","Poisson"),lty=c(1,1),lwd=c(5,4),col=c("blue","green"))
dev.off()
```

hypothesis testing of whether there is zero inflation

```{r}
thresh<-qchisq(.05/nrow(Yds),1,lower.tail=FALSE)
s<-apply(Yds,1,score_test,colMeans(Yds))
hist(s)
abline(v=thresh,lty=2)
table(s>thresh)
l<-apply(Yds,1,lr_test,colMeans(Yds))
hist(l)
abline(v=thresh,lty=2)
table(l>thresh)

pd$zero_inflated<-ifelse(l>thresh,"zero_inflated","normal")
#predict_zeros<-function(x){exp(-exp(x))} #poisson
predict_zeros<-function(x){(1-exp(x)/N)^N} #binomial
ggplot(pd)+stat_function(data=data.frame(x=c(min(pd$log_mean),max(pd$log_mean))),aes(x),fun=predict_zeros)+xlab("log of mean expression")+ylab("fraction of zero droplets")+geom_point(aes(x=log_mean,y=frac_zero))#,colour=zero_inflated))+scale_color_manual(breaks=c("normal","zero_inflated"),values=c("blue","red"))
#ggs("logmean_pzero_binom_ercc.pdf")
```

We rejected the null hypothesis of no zero inflation for several ERCCs (red points in the above graph). This is because the large sample size (number of cells) allows even small differences in predicted vs actual number of zeros to be "significant". None of the "significantly zero inflated" ERCCs deviate from the Poisson prediction by more than about 0.1.

```{r}
zi<-which.max(s)
(actual<-mean(Yds[zi,]==0))
(theor<-exp(-mean(Yds[zi,])))
```

```{r}
sessionInfo()
```