---
title: "Zheng ERCC Exploratory"
author: "Will Townes"
date: "12/04/2018"
output: html_document
---

ERCC spike ins

```{r}
library(SingleCellExperiment)
library(cowplot)
#library(ggplot2); theme_set(theme_bw())
library(Matrix)
source("./util/functions.R")
source("./algs/existing.R")
source("./algs/glmpca.R")
source("./util/clustering.R")
fp<-file.path
bp<-"./real/zheng_2017_ercc"
pth<-fp(bp,"results/fig")
if(!dir.exists(pth)){
  dir.create(pth,recursive=TRUE)
}
sp<-TRUE #save plots?
ggs<-function(plt,w=6,h=4,...){
  if(sp){ ggsave(file=fp(pth,plt),width=w,height=h,...) }
}
```

Load UMI counts matrix and read counts matrix

```{r}
sce<-readRDS(fp(bp,"data/01_sce_all_genes_all_cells.rds"))
m<-assay(sce,"counts")
Y<-as.matrix(m)
sz<-colSums(Y)
Ycpm<-1e6*t(t(Y)/sz)
Yl2<-log2(1+Ycpm)
z<-log10(sz)
pz<-colMeans(Y==0)
Ypr<-null_residuals(Y,type="pearson")
Ydr<-null_residuals(Y,type="deviance")
rc<-assay(sce,"read_counts")
```

plot histogram of a particular ERCC for reads and umis

```{r}
rmx<-apply(Y,1,max)
g<-match(20,rmx)
gn<-rownames(Y)[g]
pd<-list(data.frame(prot="raw_umi_counts",expr=Y[g,]))
pd[[2]]<-data.frame(prot="raw_read_counts",expr=rc[g,])
pd<-do.call(rbind,pd)
plts<-list()
ymax<-250
(plts[[1]]<-ggplot(subset(pd,prot=="raw_umi_counts"),aes(x=expr))+geom_bar()+ylim(NA,ymax)+theme_bw()+xlab(paste(gn,"UMI counts"))+ylab("number of droplets in bin"))
ggs("ercc_umi_distr.pdf")
(plts[[2]]<-ggplot(subset(pd,prot=="raw_read_counts"),aes(x=expr))+geom_histogram(bins=100)+ylim(NA,ymax)+theme_bw()+xlab(paste(gn,"read counts"))+ylab("number of droplets in bin"))#+ylab(NULL))
ggs("ercc_read_distr.pdf")
plot_grid(plotlist=plts)
ggs("reads_vs_umis_ercc.pdf")
```

Fraction of zeros in each droplet determined by total UMIs.

```{r}
#ggplot(data.frame(z=z,pz=pz),aes(x=z,y=pz))+geom_point()+theme_bw()+xlab("log10(total UMI per droplet)")+ylab("fraction of zeros per droplet")
#ggs("total_umi_vs_pzero_ercc.pdf")

pd<-data.frame(sz=sz,pz=pz)
(plt<-ggplot(pd,aes(x=sz,y=pz))+geom_point()+theme_bw()+xlab("total UMI per droplet")+ylab("fraction of zeros per droplet")+scale_x_log10())
(plt2<-ggExtra::ggMarginal(plt,type="histogram",fill="white",bins=100))
ggs("total_umi_vs_pzero_ercc.pdf",plot=plt2)
```

Log transformation distorts distribution

```{r}
#find gene where max count across all droplets was 20
rmx<-apply(Y,1,max)
g<-match(20,rmx)
gn<-rownames(Y)[g] #ERCC-00163
plts<-list()
(plts[["umi"]]<-ggplot(data.frame(x=Y[g,]),aes(x=x))+geom_bar()+theme_bw()+xlab("counts")+ylab(NULL))
(plts[["cpm"]]<-ggplot(data.frame(x=Ycpm[g,]),aes(x=x))+geom_histogram(bins=100)+theme_bw()+xlab("CPM")+ylab(NULL))
(plts[["logcpm"]]<-ggplot(data.frame(x=Yl2[g,]),aes(x=x))+geom_histogram(bins=100)+theme_bw()+xlab("log2(1+CPM)")+ylab(NULL))
(plts[["logcptt"]]<-ggplot(data.frame(x=log2(1+Ycpm[g,]/100)),aes(x=x))+geom_histogram(bins=100)+theme_bw()+xlab("log2(1+CPTT)")+ylab(NULL))
plot_grid(plotlist=plts,nrow=2)
ggs("counts_vs_log_cpm_histograms_ercc.pdf",w=7.5,h=5)
```

Library Size/ Fraction of Zeros dominates first PC even after normalization/ log transform

```{r}
factors<-pca(Ycpm,2)
pd<-cbind(factors,zero_frac=pz,log_total_umi=z)
round(cor(pd[,colnames(pd)!="cl"]),1)

factors<-pca(Yl2,10)
pd<-cbind(factors[,1:5],zero_frac=pz,log_total_umi=z)
round(cor(pd[,colnames(pd)!="cl"]),1)
if(cor(pd$zero_frac,pd$dim1)<0) pd$dim1<- -pd$dim1
ggplot(pd,aes(x=dim1,y=dim2,colour=zero_frac))+geom_point()+scale_color_continuous(low="blue",high="red")
ggs("log_cpm_pca12_zerofrac_ercc.pdf")
ggplot(pd,aes(x=zero_frac,y=dim1,colour=log_total_umi))+geom_point()+scale_color_continuous(low="red",high="blue")+xlab("zero fraction")+ylab("dimension 1")
ggs("log_cpm_pca1_zerofrac_total_umi_ercc.pdf")


#apply clustering to the result of PCA
cl<-seurat_cluster(factors,res=0.5)
pd<-cbind(cl,zero_frac=pz)
ggplot(pd,aes(x=factor(cluster),y=zero_frac))+geom_boxplot()+xlab("Seurat cluster ID")+ylab("zero fraction")
ggs("pca_seurat_cluster_zero_frac_ercc.pdf")

#tsne on top 50 PCs
tsf<-tsne(Yl2,2,perplexity=30,pca=TRUE)
pd2<-cbind(tsf,zero_frac=pz,log_total_umi=z)
#nonlinear color scale from https://stackoverflow.com/a/12838299
qn<-quantile(pd2$zero_frac, seq(from=0,to=1,length.out=20))
cols<-colorRampPalette(c("blue","purple","red"))(20)
vals<-scales::rescale(qn) 
ggplot(pd2,aes(x=dim1,y=dim2,colour=zero_frac))+geom_point(size=1)+scale_color_gradientn(colours=cols,values=vals)+xlab("dimension 1")+ylab("dimension 2")
ggs("log_cpm_tsne12_zerofrac_ercc.pdf")
```

same as above but using GLM-PCA

```{r}
factors<-glmpca(Y,10,verbose=TRUE)$factors
pd<-cbind(factors,zero_frac=pz,log_total_umi=z)
if(cor(pd$zero_frac,pd$dim1)<0) pd$dim1<- -pd$dim1
ggplot(pd,aes(x=dim1,y=dim2,colour=zero_frac))+geom_point()+scale_color_continuous(low="blue",high="red")
ggs("glmpca12_zerofrac_ercc.pdf")
ggplot(pd,aes(x=zero_frac,y=dim1,colour=log_total_umi))+geom_point()+scale_color_continuous(low="red",high="blue")+xlab("zero fraction")+ylab("dimension 1")
ggs("glmpca1_zerofrac_total_umi_ercc.pdf")
round(cor(pd[,colnames(pd)!="cl"]),1)

#apply clustering to the result of GLMPCA
cl<-seurat_cluster(factors,res=0.5)
pd<-cbind(cl,zero_frac=pz)
ggplot(pd,aes(x=factor(cluster),y=zero_frac))+geom_boxplot()+xlab("Seurat cluster ID")+ylab("zero fraction")
ggs("glmpca_seurat_cluster_zero_frac_ercc.pdf")

#tsne on top 50 GLMPCA factors
ctl<-list(maxIter=500,eps=1e-4)
toc<-system.time(res<-glmpca(Y,50,verbose=TRUE,ctl=ctl)) #5mins
factors<-res$factors
tsf<-tsne(t(factors),2,perplexity=30,pca=FALSE,rmzero=FALSE)
pd2<-cbind(tsf,zero_frac=pz,log_total_umi=z)
#nonlinear color scale from https://stackoverflow.com/a/12838299
qn<-quantile(pd2$zero_frac, seq(from=0,to=1,length.out=20))
cols<-colorRampPalette(c("blue","purple","red"))(20)
vals<-scales::rescale(qn) 
ggplot(pd2,aes(x=dim1,y=dim2,colour=zero_frac))+geom_point(size=1)+scale_color_gradientn(colours=cols,values=vals)+xlab("dimension 1")+ylab("dimension 2")
ggs("glmpca_tsne12_zerofrac_ercc.pdf")
```

compare clustering of log-CPM pca vs glmpca

```{r}
silfunc<-function(f,cl,m="euclidean"){
  cl<-as.integer(cl$cluster)
  d<-dist(f,method=m)
  sil<-cluster::silhouette(cl,d)
  summary(sil)$avg.width
}
L<-2
f1<-pca(Yl2,L)
f2<-glmpca(Y,L)$factors
resolution<-0.08
c1<-seurat_cluster(f1,res=resolution)
c2<-seurat_cluster(f2,res=resolution)
silfunc(f1,c1)
silfunc(f2,c2)
table(c1$cluster)
table(c2$cluster)

k<-2
c1<-kmeans_cluster(f1,k=k)
c2<-kmeans_cluster(f2,k=k)
silfunc(f1,c1)
silfunc(f2,c2)
```

larger number of perplexities

```{r}
#plx<-c(1,2,5,10,30,100)
f<-function(p){
  factors<-tsne(Yl2,2,perplexity=p,pca=FALSE)
  cbind(factors,zero_frac=pz,log_total_umi=z,perplexity=p)
}
#pd<-do.call(rbind,lapply(plx,f))
#ggplot(pd,aes(x=dim1,y=dim2,colour=zero_frac))+geom_point(size=1)+scale_color_continuous(low="blue",high="red")+facet_wrap(~perplexity,scales="free",nrow=3)
#ggs("log_cpm_tsne12_zerofrac_ercc.pdf",w=10,h=15)

#larger number of perplexities
plx2<-c(.5,1:20,40,60,80,100,150,200)
pd2<-do.call(rbind,lapply(plx2,f))
pdf(fp(pth,"log_cpm_tsne12_zerofrac_ercc_big.pdf"))
for(p in plx2){
  print(ggplot(subset(pd2,perplexity==p),aes(x=dim1,y=dim2,colour=zero_frac))+geom_point()+scale_color_continuous(low="blue",high="red")+ggtitle(paste("perplexity=",p)))
}
dev.off()
```