---
title: "Tung model selection"
author: "Will Townes"
output: html_document
---

Tung 2016 plate based UMI dataset

```{r}
suppressPackageStartupMessages(library(SingleCellExperiment))
library(ggplot2); theme_set(theme_bw())
source("./util/functions.R")
fp<-file.path
bp<-"./real/tung_2016"

plt_dir<-fp(bp,"results/fig")
if(!dir.exists(plt_dir)){ dir.create(plt_dir,recursive=TRUE) }
sp<-TRUE #save plots?
ggs<-function(plt,w=6,h=4,...){
  if(sp){ ggsave(file=fp(plt_dir,plt),width=w,height=h,...) }
}
```

Load UMI counts matrix and read counts matrix. For model comparison on negative control data, choose a single plate from one individual to ensure no batch effects.

```{r}
sce0<-readRDS(fp(bp,"data/01_sce_all_genes_all_cells.rds"))
cm<-colData(sce0)
set.seed(205) #reproducibility
ssg<-sample.int(nrow(sce0),size=5000) #subset random genes
sce<-sce0[ssg,cm$individual=="NA19098" & cm$replicate=="r3"]
m<-as.matrix(assay(sce,"counts"))
sz<-colSums(m)
z<-log10(sz)
pz<-colMeans(m==0)
rc<-as.matrix(assay(sce,"read_counts"))
rowmins<-apply(m,1,min)
rowsd<-apply(m,1,sd)
m2<-m[rowmins==0 & rowsd>1e-12,]
rowmins<-apply(rc,1,min)
rowsd<-apply(rc,1,sd)
rc2<-rc[rowmins==0 & rowsd>1e-12,]
print(dim(m2)) #3357 genes x 57 cells
# Show there are highly expressed genes
summary(rowMeans(m)) #values range up to 100-700
```

### Compare model fit for different likelihoods

Likelihoods:
* multinomial
* Poisson
* Dirichlet-multinomial
* negative binomial
* normal
* zero inflated Poisson
* zero inflated lognormal

All likelihoods include column means as offset. Have to remove genes with no zero values in order to include zero inflated models in comparisons.

```{r}
system.time(res<-bic_all(m2))
barplot(sort(res),xlab="likelihood",ylab="BIC",main="UMI counts",log="y")
res_rc<-bic_all(rc2)
barplot(sort(res_rc),xlab="likelihood",ylab="BIC",main="read counts")
dd<-as.data.frame(res)
colnames(dd)<-"BIC"
dd$dat<-"umi_counts"
dd$lik<-rownames(dd)
dd2<-as.data.frame(res_rc)
colnames(dd2)<-"BIC"
dd2$dat<-"read_counts"
dd2$lik<-rownames(dd2)
dd<-rbind(dd,dd2)
rownames(dd)<-NULL
dd<-dd[,c(2,3,1)]
write.table(dd,fp(bp,"results/gof_bic.txt"),quote=FALSE,row.names=FALSE)
```

The best performing likelihood is multinomial followed by Poisson and Dirichlet-multinomial for UMI counts.

```{r}
pd<-read.table(fp(bp,"results/gof_bic.txt"),header=TRUE)
levels(pd$lik)[levels(pd$lik)=="normal"]<-"nml"
plt_func<-function(d){
  d$lik<-factor(d$lik,levels=d$lik[order(d$BIC)])
  ggplot(d,aes(x=lik,y=BIC))+geom_point(size=3)+xlab("likelihood")
}
plt_func(subset(pd,dat=="umi_counts"))
ggs("gof_bic_tung_umi.pdf",w=3,h=2)
plt_func(subset(pd,dat=="read_counts"))
ggs("gof_bic_tung_reads.pdf",w=3,h=2)
```
