---
title: "Gene Filtering Simulation"
author: "Will Townes"
output: html_document
---

```{r}
library(tidyverse); theme_set(theme_bw())
#suppressPackageStartupMessages(library(SingleCellExperiment))
source("./util/functions_genefilter.R")
```

We want to be sure feature selection by deviance is not just finding highly expressed genes. In particular, if there are genes with high but constant expression, we want to make sure they are not picked up by deviance filtering. We will create three types of genes:
1. genes with low expression 
2. genes with high but constant expression
3. genes with high expression but variable expression (differentially expressed).

We will make sure deviance finds only the third type of gene, whereas we expect highly expressed genes to find both types 2 and 3.

```{r}
set.seed(101)
N<-500; J<-5000 #number of cells and genes
sz<-rgamma(N,shape=10,rate=10)
summary(sz)
mu1<-.1; mu2<-5; mu3a<-1; mu3b<-9
Y1<-replicate(J/2,rnbinom(N,size=.1,mu=mu1*sz))
Y2<-replicate(J/4,rpois(N,mu2*sz))
Y3a<-replicate(J/4,rnbinom(N/2,size=.1,mu=mu3a*sz[1:(N/2)]))
Y3b<-replicate(J/4,rnbinom(N/2,size=.1,mu=mu3b*sz[((N/2)+1):N]))
Y3<-rbind(Y3a,Y3b)
Y<-t(cbind(Y1,Y2,Y3))
summary(colSums(Y))
summary(1-colMeans(Y>0))
max(Y)
gtype<-c(rep("low",J/2),rep(c("high_const","high_var"),each=J/4))
rm<-rowMeans(Y)
summary(rm)
boxplot(rm~gtype) #ensure high_var and high_const have same row means
gm<-data.frame(gtype=gtype)
cm<-data.frame(celltype=rep(c("a","b"),each=N/2),nUMI=colSums(Y))
```

alternative approach with low constant and low variable genes.

```{r}
set.seed(101)
#number of cells and genes, N should be divisible by 2, J should be divisible by 100
N<-500; J<-5000 
sz<-rgamma(N,shape=10,rate=10)
summary(sz)
refmu<-data.frame(gtype=c("low-const","high-const","low-var","high-var"),a=c(.2,2,.1,1),b=c(.2,2,.3,3))
nbtheta<-.1
Y1<-replicate(.8*J,rpois(N,refmu$a[1]*sz)) #low, constant
Y2<-replicate(.1*J,rpois(N,refmu$a[2]*sz)) #high, constant
mu3<-rep(c(refmu$a[3],refmu$b[3]),each=N/2)
Y3<-replicate(.05*J,rnbinom(N,size=nbtheta,mu=mu3*sz)) #low, variable
#Y3b<-replicate(.05*J,rnbinom(N/2,size=nbtheta,mu=refmu$b[3]*sz[((N/2)+1):N])) #low, variable group b
#Y3<-rbind(Y3a,Y3b)
mu4<-rep(c(refmu$a[4],refmu$b[4]),each=N/2)
Y4<-replicate(.05*J,rnbinom(N,size=nbtheta,mu=mu4*sz)) #high, variable
#Y4b<-replicate(J/4,rnbinom(N/2,size=nbtheta,mu=refmu$b[3]*sz[((N/2)+1):N])) #low, variable group b
Y<-t(cbind(Y1,Y2,Y3,Y4))
summary(colSums(Y))
summary(1-colMeans(Y>0))
max(Y)
gtype<-c(rep("low_const",.8*J),rep("high_const",.1*J),rep(c("low_var","high_var"),each=.05*J))
gm<-data.frame(gtype=gtype,rm=rowMeans(Y))
summary(gm$rm)
#ensure high_var and high_const have same row means
ggplot(gm,aes(x=gtype,y=rm))+geom_boxplot()+scale_y_log10()+xlab("gene type")+ylab("average expression")
```

```{r}
sce<-SingleCellExperiment(assays=list(counts=Y),colData=cm,rowData=gm)
rownames(sce)<-paste0("gene",1:J)
colnames(sce)<-paste0("cell",1:N)
cl<-scran::quickCluster(sce)
table(cm$celltype,cl)
sce<-scran::computeSumFactors(sce,clusters=cl)
sce<-scater::normalize(sce)
rk<-rank_all_genes(sce)
cor(rk)
```

```{r}
pd<-cbind(rk,genetype=gtype)
pd2<-pd %>% gather(key="genefilter",value="rank",-genetype)
ggplot(pd2,aes(x=genefilter,y=rank,fill=genetype))+geom_boxplot()+scale_y_reverse()+xlab("feature selection method")+ylab("informativeness rank")
ggsave("./simulations/genefilter_sim_boxplot.pdf",width=6,height=4)
```

make contingency tables

```{r}
rk2<-as.data.frame(rk<=1000)
table(rk2[,c(1,3,2)])
```

pairwise comparisons

Comparison 1: gene A with high, constant expression, gene B with moderate but variable expression

Comparison 2: Both genes with same mean and variance, gene A is bimodal and gene B is unimodal

Comparison 3: Both genes with same mean and variance, gene A with zero inflation and gene B without zero inflation

```{r}
N<-1000
Numi<-10000
bd<-function(x){mu<-mean(x); binomial_deviance(x,mu/Numi,Numi) }
getstats<-function(x){
  mu<-mean(x)
  v<-var(x)
  c(mean=mu,var=v,fano=v/mu,dev=bd(x))
}

set.seed(101)
a<-data.frame(gene="a",counts=rpois(N,5))
b<-data.frame(gene="b",counts=rnbinom(N,size=2,mu=3))
ggplot(rbind(a,b),aes(x=counts,fill=gene))+geom_bar(position="dodge")
res<-cbind(a=getstats(a$counts),b=getstats(b$counts))
signif(res,2)

set.seed(101)
a<-data.frame(gene="a",counts=rpois(N,rep(c(1.1,6.9),each=N/2)))
b<-data.frame(gene="b",counts=rnbinom(N,size=1.8,mu=4))
ggplot(rbind(a,b),aes(x=counts,fill=gene))+geom_histogram(bins=50)+facet_grid(gene~.)
res<-cbind(a=getstats(a$counts),b=getstats(b$counts))
signif(res,2)

set.seed(101)
pzero<-0.3; p<-1-pzero; lambda<-5
z<-rbinom(N,1,p); x<-rpois(N,lambda)
a<-data.frame(gene="a",counts=z*x)
b<-data.frame(gene="b",counts=rnbinom(N,size=.9*p/(1-p),mu=1.1*lambda*p))
ggplot(rbind(a,b),aes(x=counts,fill=gene))+geom_bar(position="dodge")#+geom_histogram(bins=50)+facet_grid(gene~.)
res<-cbind(a=getstats(a$counts),b=getstats(b$counts))
signif(res,2)
```
